Smart Representations: {R}ationality, Regret, Evolution \& the Environment

### 2015-11-06 paper/paper.tex
minor edits

Evolutionary game theory has become an established philosophical tool for probing into
conceptual issues whose complexity requires mathematical modeling. However, most applications
look at evolutionary selection of behavior in a single \emph{stage game}. There are two common
simplifications in this approach. Firstly, the stage game is usually considered fixed, closed
and immutable. Secondly, what evolves are behavioral strategies for the stage game. It has
consequently been argued that, instead of focusing on a fixed, immutable stage game and the
behavioral selection for that stage game alone, we should rather study the evolutionary
competition of \emph{choice mechanisms} under selective pressure in a rich and variable
environment \citep[e.g.][]{FawcettHamblin2013:Exposing-the-be,McNamara2013:Towards-a-Riche}.
But only little attention has been paid to modeling the evolutionary competition between choice
mechanisms in a game-theoretic framework \citep[but
see][]{ZollmanSmead2010:Plasticity-and-,SmeadZollman2013:The-Stability-o}.\todo{More references
  Harley 1981!!!, O Conor \dots}.

If agents deal with a rich and varied environment, they face many different choice
situations. A crucial part of a choice mechanism (conceived of as a mapping from a choice
situation to a concrete choice of action) is a \emph{subjective representation} of that choice
situation, i.e., the agents' manner of representing their preferences and the manner of forming
decision relevant beliefs. Since almost any reasonable learning or decision-making process will
make use of some (subjective, agent-internal (possibly unconscious)) representation of the
quality of choice options, the question of which such representations are ecologically valuable
(i.e., lead to high fitness) is central and has been subject of recent interest
(\textcolor{red}{CITE literature on evolution of preferences}). Na\"ive intuition might
convince us that preferences that veridically represent the actual, true fitness measure that
evolutionary selection is based upon must surely be the best. We demonstrate that, if we take
into account how agents represent uncertainty, this na\"ive conclusion is premature. Under
certain reasonable epistemic conditions, non-veridical preference representations in terms of
regret can be evolutionarily beneficial.

To show this, we study the evolutionary competition of choice mechanisms, or more specifically
the competition of subjective representations of choice situations, in a variable
environment. Concretely, we look at evolutionary ``meta-games'' whose ``acts'' are types of
subjective representations of games and whose ``payoffs'' represent the average expected
fitness that choice mechanisms using these representation types would accrue when playing
arbitrary games (from a given class, with a given occurrence probability) against other
subjective representations of games. In this sense, a meta-game captures (the modeller's)
assumptions about the relevant statistics of the environment in which evolutionary forces
operate, while we are still able to use standard methods from evolutionary game theory, like
stability or evolutionary dynamics.

@@

For the purposes of this paper, we think of a choice rule as a function that takes a
(strategic) game as input and outputs a choice of action for that game. The literatures has
suggested a number of choice rules that aspire to capture a normative element of rational
choice. For reasons that will become clear as we go along, we focus here on three classical
choice rules: (i) maximin, (ii) Laplace and (iii) regret minimization.

@@

A \emph{choice mechanism} is a map from choice situations to action choices that, unlike choice
rules, contains an explicit level of subjective representation of a decision situation. For our
purposes, we should think of a choice mechanism is a pair of functions $\gamma$ and $\kappa$,
where $\gamma$ is a \emph{subjective representation scheme} and $\kappa$ is an \emph{action
  selection function} (see FigureXYZ).\todo{make figure when notation and terminology are
  fixed}\ The subjective representation scheme $\gamma$ maps a game (or more generally any
decision situation) onto a subjective representation of the game (or choice situation) and
other decision-relevant beliefs (e.g., about what the opponent is likely to play). The action
selection function $\kappa$ takes such a subjective representation that $\gamma$ returns, and
maps it onto a concrete action choice. The crucial bit for our purposes here is the
representation function $\gamma$, because we are interested in which ways of conceptualizing a
decision situation are evolutionarily successful.

More generally even, a choice mechanism could be a set of distinct heuristics (the DM's
\emph{adaptive toolbox}) that have little in common
\citep[e.g.][]{TverskyKahnemann1981:The-Framing-of-,GigerenzerGoldstein1996:Reasoning-the-F,ScheibehenneRieskamp2013:Testing-the-Ada}. But
to keep matters simple and implementable for a start, we will here only look at choice
mechanisms that differ only with respect to their subjective representation function $\gamma$,
which in turn is conceived of as a \emph{uniform} way of forming preference and belief
\emph{types}.

Formally, given a class $ \mathcal{G} $ of games, $\gamma: \mathcal{G} \rightarrow \Theta \times E$ is a function that assigns a subjective preference $\theta \in \Theta$ and an epistemic state (or belief type) $e \in E$ to any game $G \in \mathcal{G}$.\footnote{Given a game $G$, the set $ \Theta $ is the set of all subjective preferences, or subjective utility functions, on $G$. I.e., $\Theta$ is the set of all functions $\theta: A \rightarrow \mathbb{R}$.} 
Then, given a set of subjective
  preferences $ \Theta $ and
  a set of or belief types $E$, an action selection function is a function
  $\kappa: \Theta \times E \rightarrow A_i $, that associates any subjective representation of the game $ (\theta,e) \in \Theta \times E$ with an action $a^{*} \in A_i $.

More concretely, we focus on two (\textit{subjective}) \emph{preference types} and two \emph{belief
    types}. We view a preference type as a function $\tau: \mathcal{G} \rightarrow \Theta$, and
  a belief type as a general disposition to form beliefs about the co-player’s behavior. The
  two possible dispositions that we take account of are precise probabilistic belief and
  imprecise belief, as we will explain in the next section.

Since we are most interested here in studying whether evolutionary pressure could
  support non-veridical subjective preferences, the two preference types that we consider are: the veridical type $\tau^\pi$, who uses the actual payoff function
  $\pi$ of the actual game $G$, and the regret type $\tau^{reg}$, who looks at the regret
  transformation of the actual game $G$.

Finally, the action selection function $\kappa$ that we use in this paper is \textit{maxmin expected
    utility} [GS89], for reasons that will be clear and justified in the following two
  sections.

% It turns out that whether $\tau^{reg}$ can have a fitness advantage over $\tau^\pi$ depends on
% the general \emph{kind} of (probabilistic) beliefs that DM's form about the opponent's
% behavior. In other words, we even abstract away from the content of the DM's beliefs about his
% opponents' behavior that he could have acquired by learning or observation, but first turn on
% the more general question of how uncertainty should be represented in order to interact with
% subjective preferences in such a way that the resulting behavior will be fitness enhancing for
% the DM in the long run. We consider therefore two \emph{belief types} (to be defined below):
% imprecise belief types $\Delta(A_{-i})$, who have a set of probabilistic beliefs about the
% opponents' behavior and precise belief types $\overline{\mu}$ who have a single probabilistic
% belief.

@@

In evolutionary game theory, for instance, players in a population obviously face uncertainty about the composition of the population that they are part of, and consequently about the co-player that they are randomly paired with at each round and about the co-player's action. We represent this uncertainty in two different ways that correspond to the two approaches discussed above: a Bayesian uniform distribution and an imprecise probability set. This amounts to having two different belief types, $ \overline{\mu}$ and $ \Delta(A_{-i})$, where $\overline{\mu}$ stands for the uniform distribution and $\Delta(A_{-i})$ is the probability set, defined as follows:

@@

\item $\overline{\mu} \in \Delta(A_{-i})$ is the probability measure over the co-player's actions such that: $\forall a_{-i} \in A_{-i}$, $\overline{\mu}(a_{-i})= \frac{1}{|A_{-i}|}$.

>>>>

Evolutionary game theory has become an established philosophical tool for probing into
conceptual issues whose complexity requires mathematical modeling. However, most applications
look at evolutionary selection of behavior in a single \emph{stage game}. There are two common
simplifications in this approach. Firstly, the stage game is usually considered fixed, closed
and immutable. Secondly, what evolves are behavioral strategies for the stage game. It has
consequently been argued that, instead of focusing on a fixed, immutable stage game and the
behavioral selection for that stage game alone, we should rather study the evolutionary
competition of \emph{choice mechanisms} under selective pressure in a rich and variable
environment \citep[e.g.][]{FawcettHamblin2013:Exposing-the-be,McNamara2013:Towards-a-Riche}.
But only little attention has been paid to modeling the evolutionary competition between choice
mechanisms in a game-theoretic framework \citep[but
see][]{ZollmanSmead2010:Plasticity-and-,SmeadZollman2013:The-Stability-o}.\todo{More references
  Harley 1981!!!, O Conor \dots}

If agents deal with a rich and varied environment, they face many different choice
situations. A crucial part of any choice mechanism is a \emph{subjective representation} of the
choice situation, e.g., the agent's manner of representing their preferences and her way of
forming decision relevant beliefs in the case of uncertainty. Since almost any reasonable
learning or decision-making process will make use of some (subjective, agent-internal (possibly
unconscious)) representation of the quality of choice options, the question of which such
representations are ecologically valuable (i.e., lead to high fitness) is central and has been
subject of recent interest (\textcolor{red}{CITE literature on evolution of
  preferences}). Na\"ive intuition might convince us that preferences that veridically
represent the actual, true fitness measure that evolutionary selection is based upon must
surely be the best. We demonstrate that, if we take into account how agents represent
uncertainty, this na\"ive conclusion is premature. Under certain reasonable epistemic
conditions, non-veridical preference representations can be evolutionarily beneficial.

To show this, we study the evolutionary competition of choice mechanisms, or more specifically
the competition of subjective representations of choice situations, in a variable
environment. Concretely, we look at evolutionary ``meta-games'' whose ``acts'' are types of
subjective representations of games and whose ``payoffs'' represent the average expected
fitness that choice mechanisms using these representation types would accrue when playing
arbitrary games (from a given class, with a given occurrence probability). In this sense, a
meta-game captures (the modeller's) assumptions about the relevant statistics of the
environment in which evolutionary forces operate, while we are still able to use standard
methods from evolutionary game theory, like stability or evolutionary dynamics.

@@

For the purposes of this paper, we think of a choice rule as a function that takes a
(strategic) game as input and outputs a choice of action for that game. The literature has
suggested a number of choice rules that aspire to capture a normative element of rational
choice. We focus here on three classical choice rules: (i) maximin, (ii) Laplace and (iii)
regret minimization.

@@

A \emph{choice mechanism} is a map from choice situations to action choices that, unlike choice
rules, contains an explicit level of subjective representation of a decision situation. For our
purposes, we should think of a choice mechanism as a pair of functions $\gamma$ and $\kappa$,
where $\gamma$ is a \emph{subjective representation scheme} and $\kappa$ is an \emph{action
  selection function} (see FigureXYZ).\todo{make figure when notation and terminology are
  fixed}\ The subjective representation scheme $\gamma$ maps a game (or more generally any
decision situation) onto a subjective representation of the game (or choice situation) and
other decision-relevant beliefs (e.g., about what the opponent is likely to play). The action
selection function $\kappa$ takes such a subjective representation that $\gamma$ returns and
maps it onto a concrete action choice. The crucial bit for our purposes here is the
representation function $\gamma$, because we are interested in studying which ways of
conceptualizing a decision situation are evolutionarily successful.

A choice mechanism could be a set of distinct heuristics (the DM's \emph{adaptive toolbox})
that have little in common
\citep[e.g.][]{TverskyKahnemann1981:The-Framing-of-,GigerenzerGoldstein1996:Reasoning-the-F,ScheibehenneRieskamp2013:Testing-the-Ada}. But
to keep matters simple and implementable for a start, we will here only look at choice
mechanisms that differ only with respect to their subjective representation function $\gamma$,
which in turn is conceived of as a \emph{uniform} way of forming preference and belief
\emph{types}: we look at general \emph{subjective representation schemes}.

Formally, given a class $ \mathcal{G} $ of strategic games,
$\gamma_i: \mathcal{G} \rightarrow \Theta_i \times E_i$ is a function that assigns a subjective
preference $\theta_i \in \Theta_i$ and an epistemic state $e_i \in E_i$ to any game
$G \in \mathcal{G}$ for some agent $i$. A subjective preference $\theta_i$ for game $G$ and
agent $i$ is just a payoff function for $i$ suitable to the acts and co-players in
$G$. Crucially, this subjective payoff function can be different from the actual payoff in $G$,
which we interpret as the objective fitness. An epistemic state $e_i$ for game $G$ and agent
$i$ is some specification of $i$'s uncertainty about the co-players' behavior. Finally, an
action selection function for agent $i$ is a function
$\kappa_i: \Theta_i \times E_i \rightarrow \pow{A_i} $, that associates any subjective
representation of the game $ (\theta_i,e_i) \in \Theta_i \times E_i$ with a set of prescribed
actions for $i$ in $G$ under $i$'s subjective representation of $G$. We drop the index $i$ for
readability in the following.

We focus on two (\textit{subjective}) \emph{preference types} and two \emph{belief types}. We
view a preference type as a function $\tau: \mathcal{G} \rightarrow \Theta$, and a belief type
as a general disposition to form beliefs about the co-player’s behavior. The two possible
dispositions that we take account of are precise probabilistic belief and imprecise belief, as
we will explain in the next section. Since we are most interested here in studying whether
evolutionary pressure could support non-veridical subjective preferences, the two preference
types that we consider are: the veridical type $\tau^\pi$, who uses the actual payoff function
$\pi$ of the actual game $G$, and the regret type $\tau^{reg}$, who looks at the regret
transformation of the actual game $G$. Finally, the action selection function $\kappa$ that we
use in this paper is \textit{maxmin expected utility} [GS89], for reasons that will be clear
and justified in the following two sections.

% It turns out that whether $\tau^{reg}$ can have a fitness advantage over $\tau^\pi$ depends on
% the general \emph{kind} of (probabilistic) beliefs that DM's form about the opponent's
% behavior. In other words, we even abstract away from the content of the DM's beliefs about his
% opponents' behavior that he could have acquiresd by learning or observation, but first turn on
% the more general question of how uncertainty should be represented in order to interact with
% subjective preferences in such a way that the resulting behavior will be fitness enhancing for
% the DM in the long run. We consider therefore two \emph{belief types} (to be defined below):
% imprecise belief types $\Delta(A_{-i})$, who have a set of probabilistic beliefs about the
% opponents' behavior and precise belief types $\overline{\mu}$ who have a single probabilistic
% belief.

@@

In evolutionary game theory, for instance, players in a population obviously face uncertainty about the composition of the population that they are part of, and consequently about the co-player that they are randomly paired with at each round and about the co-players' action. We represent this uncertainty in two different ways that correspond to the two approaches discussed above: a Bayesian uniform distribution and an imprecise probability set. This amounts to having two different belief types, $ \overline{\mu}$ and $ \Delta(A_{-i})$, where $\overline{\mu}$ stands for the uniform distribution and $\Delta(A_{-i})$ is the probability set, defined as follows:

@@

\item $\overline{\mu} \in \Delta(A_{-i})$ is the probability measure over the co-players' actions such that: $\forall a_{-i} \in A_{-i}$, $\overline{\mu}(a_{-i})= \frac{1}{|A_{-i}|}$.
